F\# Data: \textnormal{Making structured data first-class citizens}

### 2015-02-26 icfp-2015/paper.tex
Adding references and related work

\usepackage{amssymb}
\usepackage{multicol}
\usepackage{xcolor}
\usepackage{graphicx}
\usepackage[fleqn]{amsmath}
\usepackage[hang,flushmargin]{footmisc} 
\usepackage{graphics} 
\usepackage{stmaryrd}
\usepackage{amsthm}
\usepackage[hyphens]{url}
\usepackage{hyperref}
\usepackage{ftnright}
\usepackage[notquote]{hanging}
\usepackage[T1]{fontenc}
\usepackage{semantic}
\usepackage{enumitem}

@@

\noindent
Although the F\# Data library has been widely adopted is one of the most downloaded F\# libraries 
it has not yet been presented in an academic form. Additional documentation and source code can 
be found at \url{http://fsharp.github.io/FSharp.Data}.

@@

Sample is captured at compile time. It can come from the web or not (cache locally if you want
but from the web has nice implications).

We dont know what to do about schema change at runtime
(for some data sources, we could do more clever things though)

In practice..
Error handling - solve issues just add another example

@@

\section{Related and future work}
\label{sec:related}
~

Statistical metric makes the safety even messier (probabilistic!)

Write API

~

F\# and Agda type providers

Kathleen Fisher's work

JSON inference stuff

LINQ and XML stuff

@@

% We recommend abbrvnat bibliography style.

\bibliographystyle{abbrvnat}

% The bibliography should be embedded for final submission.

\begin{thebibliography}{}
\softraggedright

\bibitem[Smith et~al.(2009)Smith, Jones]{smith02}
P. Q. Smith, and X. Y. Jones. ...reference text...

\end{thebibliography}

>>>>

\usepackage{amssymb}
\usepackage{multicol}
\usepackage{xcolor}
\usepackage{graphicx}
\usepackage[fleqn]{amsmath}
\usepackage[hang,flushmargin]{footmisc} 
\usepackage{graphics} 
\usepackage{stmaryrd}
\usepackage{amsthm}
\usepackage[hyphens]{url}
\usepackage{hyperref}
\usepackage{ftnright}
\usepackage[notquote]{hanging}
\usepackage[T1]{fontenc}
\usepackage{semantic}
\usepackage{enumitem}
\usepackage{flushend}

@@

\noindent
Although the F\# Data library \cite{fsharp-data} has been widely adopted is one of the most downloaded F\# libraries 
it has not yet been presented in an academic form. Additional documentation and source code can 
be found at \url{http://fsharp.github.io/FSharp.Data}.

@@

An important concern about using F\# Data in practice is the handling of schema change. When using a
type provider, the sample is captured at compile-time. If the schema changes later (so that the actual
input is no longer a subtype of the samples), the program using the type provider fails at run-time
and it is the developer's responsibility to handle the exception. However, this is the same problem that
happens when reading data using any other library.

F\# Data can help discover such errors earlier. The example in Section~\ref{sec:introduction}
points the JSON type provider to a sample using a live URL. This has the advantage that a re-compilation 
fails when the schema changes, which is an indication that the program needs to be updated to reflect the
change. If this is undesirable, it is always possible to cache the sample locally.

In general, there is no better solution for plain XML, CSV and JSON data sources. Some data sources provide 
versioning support (with meta-data about how the schema changed). For those, a type 
provider could adapt automatically, but we leave this for future.

An approach that we find useful in practice is to keep a set of samples. When the program fails at run-time
(because a service returned data in an unexpected format), the new input can be added as an additional sample,
which then shows what parts of code need to be modified.

@@

\section{Related and future work}
\label{sec:related}

The F\# Data library connects two lines of research that have been previously disconnected. The first is 
extending the type systems of programming languages to accommodate external data sources and the second
is inferring types for real-world data sources.

The type provider mechanism has been introduced in F\# \cite{fsharp-typeprov,fsharp-typeprov-ddfp}, 
added to Agda \cite{agda-tp} and used in areas such as semantic web \cite{liteq}. Our paper is novel 
in that it shows the programming language theory behind a concrete type providers.

\paragraph{Extending the type systems.} 
A number of systems integrate external data formats into a programming language. Those include 
XML \cite{xduce,xduce-ml} and databases \cite{links}. In both of these, the system either requires
the user to explicitly define the schema (using the host language) or it has an ad-hoc extension 
that reads the schema (\emph{e.g.}~from a database). LINQ \cite{linq} is more general, but relies
on code generation when importing the schema.

The work that is most similar to F\# Data is the XML and SQL integration in C$\omega$ \cite{comega-xs}.
It extends a C\# like object-oriented language with types similar to our structured types 
(including nullable types, choices with subtyping and heterogeneous collections with multiplicities).
However, C$\omega$ does not infer the types from samples and extends the type system of the host
language (rather than using a general purpose embedding mechanism).

\paragraph{Advanced type systems.}
Aside from type providers, a number of other advanced type system features could be used to
tackle the problem discussed in this paper. The Ur \cite{ur} language has a rich system for working
with records; meta-programming \cite{template-hask}, \cite{th-camlp4} and multi-stage programming \cite{multi-stage}
could be used to generate code for the provided types. However, as far as we are aware, none of these 
systems have been used to provide the same level of integration with XML, CSV and JSON.

Another approach would be to use gradual typing \cite{gradual,gradual-js} and add types for structured
data formats to existing dynamic language to check, for example, JSON manipulation in JavaScript.

\paragraph{Typing real-world data.}
The second line of research related to our work focuses on inferring structure of real-world data sets.
A recent work on JSON \cite{typing-json} infers a succinct type using MapReduce to handle large number
of samples. It fuses similar types based on a type similarity measure. This is more sophisticated than
our technique, but it would make formally specifying the safety properties (Theorem~\ref{thm:safety}) difficult.
Extending our \emph{relativized safety} property to a \emph{probabilistic safety} is an interesting 
future work.

The PADS project \cite{pads-dsl,pads-ml} tackles a more general problem of handling \emph{any} data format.
The schema definitions in PADS are similar to our structural type. The structure inference for LearnPADS
\cite{pads-learn} infers the data format from a flat input stream. A type provider for PADS could follow
many of the patterns we explore in this paper, but formally specifying the safety property would be
challenging.

@@

\cite{age-of-web}

\bibliographystyle{abbrvnat}
\bibliography{paper}
